\specialhead{ACKNOWLEDGMENT}

First and foremost, I would like to thank my advisor
Prof. Mark S. Shephard, for his guidance throughout my tenure
as a graduate student. Prof. Shephard's invaluable insights
and dedication to rigor and excellence have helped me become
a better researcher. I am honored to have been given the
opportunity to join his research group. I would also like
to thank my co-advisor, Prof. Assad A. Oberai, for his
insightful guidance and kindness.

I would like to thank Prof. Antionette M. Maniatty,
Prof. Jason E. Hicken, and Prof. Jeffrey W. Banks for
kindly serving on my doctoral committee, for the valuable
education they've provided me through their courses,
and for their helpfulness and willingness to answer questions
that arose during my time as a graduate student.

The Scientific Computation Research Center provided a
great working environment and I would like to thank its
members. Max Bloomfield provided me with valuable guidance
in both research and life during my time at RPI.
I am grateful for the friendship and support of fellow
graduate students Cameron Smith, Daniel Ibanez, Zhen Li,
Li Dong, Alp Dener, Jared Crean, and Alvin Zhang.

I owe the creation of my career to Rod Douglass, who
introduced me to the world of scientific computing while
mentoring me at Los Alamos National Laboratory.
Thanks, Rod.

Finally, I would like to extend my love and gratitude
to my family: Kim, Howard, Rachel, and Noel. I would not
be where I am today without their love and support.
