%%% INTRODUCTION AND BACKGROUND
\chapter{INTRODUCTION AND BACKGROUND}
\label{chap:intro}

\let\thefootnote\relax\footnotetext{
Portions of this chapter have been submitted to:
B.~N. Granzow, A.~A. Oberai, and M.~S. Shephard,
``An automated approach for parallel adjoint-based error
estimation and mesh adaptation,'' submitted for publication.}

%%% INTRODUCTION
\section{Introduction}

Numerical simulation has become ubiquitous in engineering and scientific
practice due to the continuing increase in power and accessibility of
computational resources. For scientific and engineering applications, ensuring
the accuracy and reliability of the computed numerical solution is of primary
importance. For example, it is often necessary to design structural components
for which the von-Mises stress in the component's domain is less than a given
material yield strength when subjected to a variety of loading conditions.

In the context of finite element methods, \emph{a posteriori} error estimation
and mesh adaptation provide useful tools for approximating and controlling
the discretization error when solving partial differential equations
(PDEs) (\emph{c.f} \cite{ainsworth2011posteriori, gratsch2005posteriori,
verfurth1994adaptive}).

In an abstract finite element setting, one seeks to solve the variational
problem: find $u \in \S$ such that
%
%% aut_abstract_variational
\begin{gather}
\R(w; u) = 0 \quad \forall w \in \V,
\end{gather}
%
where $\R: \S \times \V \to \mathbb{R}$ is a semilinear form, linear in its
first argument and potentially nonlinear in its second, and $\S$ and
$\V$ are Hilbert spaces. Given the exact solution $u$ to this problem and
a corresponding finite element approximation $u^H$, the discretization error
is defined as $e := u -u^H$. Traditional \emph{a posteriori} error estimates
attempt to bound or approximate the discretization error in a global norm
$\| e \|$, such as the energy norm induced by the underlying partial
differential operator. In the past twenty years, though, adjoint-based
techniques have been developed and utilized to obtain approximations $\eta$
or approximate bounds $\hat{\eta}$ for the error $|J(u) - J(u^H)|$ in some
physically meaningful functional $J : \V \to \mathbb{R}$, such that
%
%% aut_functional_bounds_abstract
\begin{gather}
|J(u) - J(u^H)| \approx \eta < \hat{\eta}.
\end{gather}

The use of adjoint (or duality) techniques for \emph{a posteriori} error
estimation can be traced back to Babu\v{s}ka and Miller
\cite{babuska1984post1, babuska1984post2, babuska1984post3} who explored the
post-processing of functional quantities. These ideas were then expanded to the
context of adaptive finite element methods by Johnson et al.
\cite{eriksson1996computational}. Becker and Rannacher \cite{becker2001optimal}
developed an approach to functional \emph{a posteriori} error estimation for
Galerkin finite element methods called the \emph{dual weighted residual}
method. Giles and Pierce \cite{giles2003adjoint} developed an approach
conceptually similar to the dual weighted residual method, but additionally
concerned themselves with discretizations that lack Galerkin orthogonality,
such as Godunov finite volume methods. Venditti and Darmofal
\cite{venditti2000adjoint, venditti2002adjoint, venditti2003adjoint} developed
adjoint-based error estimates for arbitrary discretizations using discrete
adjoint equations based on two-level discretization schemes.
Prudhomme and Oden \cite{prudhomme1999goal, oden2001goal} developed
adjoint techniques to determine guaranteed upper and lower bounds for
linear functionals in the context of linear variational problems.

Adjoint-based error estimation and mesh adaptation have been heavily adopted
by the computational fluid dynamics (CFD) community \cite{fidkowski2011review}.
This can in part be explained by the fact that the current increase in
computing power alone may not be sufficient to accurately resolve
quantities of interest (QoIs) in CFD applications, even when very finely
generated \emph{a priori} meshes are used \cite{fidkowski2011review}. However,
despite its popularity in CFD, adjoint-based error estimation is not regularly
applied to solid mechanics applications.

In the context of solid mechanics, adaptive adjoint-based error estimation
has been used to study linear elasticity in two
\cite{rannacher1997feed, stein2007error, gonzalez2014mesh} and three
\cite{ghorashi2014goal} dimensions, two
\cite{rannacher1998posteriori, rannacher1999posteriori} and three
\cite{ghorashi2017goal} dimensional elasto-plasticity, two dimensional
thermoelasticity \cite{rabizadeh2015adaptive}, two dimensional
nonlinear elasticity \cite{larsson2002strategies}, and two dimensional
hyperelasticity \cite{whiteley2014error}. We remark that in the majority of
this literature, mesh adaptation is performed with structured adaptive mesh
refinement using quadrilateral or hexahedral elements and without any
discussion of parallelization. Additionally, none of these investigations
apply the current state of the art capabilities of parallel three-dimensional
unstructured mesh adaptation, as is currently the norm for CFD
applications.

Perhaps one reason for this discrepancy is the fact that adjoint-based error
estimation requires the development and implementation of a number of
non-trivial steps. From a high-level, the following steps must be carried out
to perform an adaptive adjoint-based error analysis:
%
\begin{enumerate}
\item Solve the original (primal) PDE of interest.
\item Using the primal solution, derive and solve an auxiliary adjoint PDE.
\item Enrich the solution to the adjoint PDE in some manner.
\item Compute error estimates using an adjoint-weighted residual method.
\item Localize the error to contributions at the mesh entity level.
\item Adapt the mesh based on the local error contributions.
\end{enumerate}
%
Additionally, modern solid mechanics applications necessitate the use of parallel
analysis, meaning each of these steps must execute effectively and
efficiently on parallel machines.

Further, both the primal residual and functional QoI can be highly nonlinear
in solid mechanics. Solving the primal and adjoint problems
requires the linearization of the primal residual and the functional QoI
with respect to the degrees of freedom of the problem. The derivation and
numerical implementation of these linearizations can be time consuming and
error-prone. As a result, adjoint-based error control has typically remained
a tool for expert analysis with a high barrier of entry.

As a final observation, we note that unstructured mesh generation and
adaptation is robust, reliable, and scalable for simplical elements. However,
for solid mechanics applications with incompressibility constraints, such as
isochoric plasticity or incompressible hyperelasticity, standard
displacement-based Galerkin finite element discretizations are known to be
unstable when using simplical elements. This fact may have further hindered
the adoption of unstructured mesh adaptation for previous adaptive
adjoint-based solid mechanics applications.

%%% OUTLINE
\section{Outline}

In order to make adjoint-based error estimation and mesh adaptation
more accessible to practitioners of solid mechanics, we have developed
a fully automated approach that addresses the difficulties outlined
above. Specifically, we have developed a \texttt{C++} application called
Goal that automates steps 1-6 based solely on the inputs of a
semilinear form $\R$ and a functional QoI $J$ for both Galerkin and
stabilized finite element methods. Chapter \ref{chap:automated}
provides a review of the mathematical background required for
adjoint-based error estimation and discusses in detail the approach
we have developed to automate the process. In addition, Chapter
\ref{chap:automated} provides examples that demonstrate this approach
executes effectively on parallel machines.

Chapter \ref{chap:mech} presents the governing equations for
finite deformation elasticity and discusses the derivation
of a stabilized mixed pressure-displacement finite element
formulation for these equations. In this chapter, the automated
approach developed in Chapter \ref{chap:automated} is discussed
specifically in the context of nonlinear elasticity, and is utilized
to investigate several examples in incompressible elasticity.
In particular, the automated approach is shown to yield both
effective error estimates in this context. Further, the ability of the
Goal application to handle mesh adaptation in the presence of complex
geometries is demonstrated by investigating a model problem arising
from the study of a biological cell embedded in an extracellular
matrix.

In Chapter \ref{chap:refine}, the automated approach presented in
Chapter \ref{chap:automated} is extended to investigate two
novel approaches for solving the adjoint problem.
These approaches proceed by solving the adjoint problem on a
nested mesh obtained via \emph{non-uniform} refinement so as to
reduce the total number of degrees of freedom for the adjoint problem
when compared to solving the adjoint problem on a mesh obtained
with \emph{uniform} refinement. However, these two approaches still
maintain a physically meaningful global enrichment of the adjoint
solution. This investigation demonstrates a key feature of the Goal
software, in that it is flexible enough to be used to conduct novel
research that falls outside of its original intended scope.

Finally, we investigate adjoint-based error estimation and mesh
adaptation for variational multiscale (VMS) methods in Chapter \ref{chap:vms}.
The VMS method provides a systematic approach to develop and derive
stabilized finite element formulations for PDEs. In this chapter,
we propose a novel strategy for enriching the solution to the
adjoint problem based on VMS techniques. We use this enrichment to
derive an adjoint-based error estimate for functional QoIs.
In this chapter, we consider the advection-diffusion equation as a
model problem with the intent that this approach will soon
be implemented in the Goal application and applied to
solid mechanics applications.

%%% CONTRIBUTIONS
\section{Contributions}

This thesis centers around contributions made to
perform and investigate adjoint-based error estimation and mesh
adaptation while largely focusing on the area of solid mechanics.
In the performance of the work for this thesis, novel contributions
have been achieved by:
%
\begin{enumerate}
\item Developing and implementing an automated approach
for adjoint-based error estimation that executes effectively
on parallel machines.
\item Applying the automated approach to investigate
adjoint-based error estimation and mesh adaptation for finite
deformation elasticity with a stabilized finite element method.
\item Advancing the ability to perform adjoint-based error
estimation and mesh adaptation for solid mechanics applications
with complex geometries.
\item Applying the automated approach to conduct an investigation
of a non-uniform refinement approach for solving adjoint problems.
\item Developing and investigating a novel approach for adjoint-based error
estimation for variational multiscale finite element methods.
\end{enumerate}
